---
title: "R with Big Data 4: Big Data and Spark ML"
author: "Garrett Grolemund and Nathan Stephens"
date: "September 27, 2016"
output: html_notebook
---

```{r setup, include = FALSE}
library(sparklyr)
library(dplyr)
library(ggplot2)
library(titanic)
```

These are the class notes (Part 4 of 4) for *R for Big Data*, a workshop taught at *Strata + Hadoop World 2016 NYC*. The notes are saved as an R Markdown Notebook. See Part 1, *Universal Tools* to learn more about how to use R Markdown Notebooks.

# Overview

You can use `sparklyr` to fit a wide variety of machine learning algorithms in Apache Spark. This analysis compares the performance of six classification models in Apache Spark on the [Titanic](https://www.kaggle.com/c/titanic) data set.

Compare the following 6 models:

* Random forest - `ml_random_forest`
* Decision tree - `ml_decision_tree`
* Gradient boosted tree - `ml_gradient_boosted_trees`
* Logistic regression - `ml_logistic_regression`
* Multilayer perceptron (neural net) - `ml_multilayer_perceptron`
* Naive Bayes - `ml_naive_bayes`

# Load the data

Parquet is a column based data format that is also compressed. It is a format often used with Spark. Load the Titanic Parquet data into a local spark cluster. 

```{r data, message=FALSE, warning=FALSE}
# Connect to local spark cluster and load data
sc <- spark_connect(master = "local", version = "2.0.0")
spark_read_parquet(sc, name = "titanic", path = "data/titanic-parquet")
titanic_tbl <- tbl(sc, "titanic")
```

***

# Tidy the data

Tidy the data in preparation for model fitting. `sparkyr` uses `dplyr` syntax when connecting to the Spark SQL API and specific functions functions for connecting to the Spark ML API. 

## Spark SQL transforms

Use feature transforms with Spark SQL. Create new features and modify existing features with `dplyr` syntax.

1. Family_Size - Create number of siblings and parents
2. Pclass - Format passenger class as character not numeric
3. Embarked - Remove a small number of missing records
4. Age - Impute missing age with average age

```{r sparkSQL}
# Transform features with Spark SQL API
titanic2_tbl <- titanic_tbl %>% 
  mutate(Family_Size = SibSp + Parch + 1L) %>% 
  mutate(Pclass = as.character(Pclass)) %>%
  filter(!is.na(Embarked)) %>%
  mutate(Age = if_else(is.na(Age), mean(Age), Age)) %>%
  sdf_register("titanic2")
```

> Tip: `sdf_register` is used to save our table for later analysis.


## Spark ML transforms

Use feature transforms with Spark ML. Use `ft_bucketizer` to bucket family sizes into groups (1, 2, 5, 12).

```{r sparkFT}
# Transform family size with Spark ML API
titanic_final_tbl <- titanic2_tbl %>%
  mutate(Family_Size = as.numeric(Family_size)) %>%
  sdf_mutate(
    Family_Sizes = ft_bucketizer(Family_Size, splits = c(_________))
    ) %>%
  mutate(Family_Sizes = as.character(as.integer(Family_Sizes))) %>%
  sdf_register("titanic_final")
```

> Tip: You can use magrittr pipes to chain dplyr commands with sparklyr commands. For example, `mutate` is a dplyr command that accesses the Spark SQL API whereas `sdf_mutate` is a sparklyr command that accesses the Spark ML API.

## Train-validation split

Randomly partition the data into train and test sets where training gets a random 75% of the records and test gets the other 25% of the records.

```{r partition}
# Partition the data
partition <- titanic_final_tbl %>% 
  mutate(Survived = as.numeric(Survived), SibSp = as.numeric(SibSp), Parch = as.numeric(Parch)) %>%
  select(Survived, Pclass, Sex, Age, SibSp, Parch, Fare, Embarked, Family_Sizes) %>%
  sdf_partition(train = _________, test = _________, seed = 8585)

# Create table references
train_tbl <- partition$train
test_tbl <- partition$test
```

> Tip: Use `sdf_partition` to create training and testing splits.

***

# Train the models

Train multiple machine learning algorithms on the training data. Score the test data with the fitted models.

## Logistic regression

Logistic regression is one of the most common classifiers. Train the logistic regression and examine the predictors.

```{r train}
# Model survival as a function of several predictors
ml_formula <- formula(Survived ~ Pclass + Sex + Age + SibSp + Parch + Fare + Embarked + Family_Sizes)

# Train a logistic regression model
(ml_log <- ml_logistic_regression(train_tbl, ml_formula))
```

## Other ML algorithms

Run the same formula using the other machine learning algorithms. Notice that training times vary greatly between methods.

```{r ml}
## Decision Tree
ml_dt <- ml_decision_tree(train_tbl, ml_formula)

## Random Forest
ml_rf <- ml_random_forest(train_tbl, ml_formula)

## Gradient Boosted Tree
ml_gbt <- ml_gradient_boosted_trees(train_tbl, ml_formula)

## Naive Bayes
ml_nb <- ml_naive_bayes(train_tbl, ml_formula)

## Neural Network
ml_nn <- ml_multilayer_perceptron(train_tbl, ml_formula, layers = c(11,15,2))
```

## Validation data

Score the test data with the trained models.

```{r score}
# Bundle the modelss into a single list object
ml_models <- list(
  "Logistic" = ml_log,
  "Decision Tree" = ml_dt,
  "Random Forest" = ml_rf,
  "Gradient Boosted Trees" = ml_gbt,
  "Naive Bayes" = ml_nb,
  "Neural Net" = ml_nn
)

# Create a function for scoring
score_test_data <- function(model, data=test_tbl){
  pred <- sdf_predict(model, data)
  select(pred, Survived, prediction)
}

# Score all the models
ml_score <- lapply(ml_models, score_test_data)
```

***

# Compare results

Compare the model results. Examine performance metrics: lift, AUC, and accuracy. Also examine feature importance to see what features are most predictive of survival.

## Model lift

Lift compares how well the model predicts survival compared to random guessing. Use the function below to estimate model lift for each scored decile in the test data. The lift chart suggests that the tree models (random forest, gradient boosted trees, or the decision tree) will provide the best prediction.


```{r lift}
# Lift function
calculate_lift <- function(scored_data) {
  scored_data %>%
    mutate(bin = ntile(desc(prediction), 10)) %>% 
    group_by(bin) %>% 
    summarize(count = sum(Survived)) %>% 
    mutate(prop = count / sum(count)) %>% 
    arrange(bin) %>% 
    mutate(prop = cumsum(prop)) %>% 
    select(-count) %>% 
    collect() %>% 
    as.data.frame()
}

# Initialize results
ml_gains <- data.frame(bin = 1:10, prop = seq(0, 1, len = 10), model = "Base")

# Calculate lift
for(i in names(ml_score)){
  ml_gains <- ml_score[[i]] %>%
    calculate_lift %>%
    mutate(model = i) %>%
    rbind(ml_gains, .)
}

# Plot results
ggplot(ml_gains, aes(x = bin, y = prop, colour = model)) +
  geom_point() + geom_line() +
  ggtitle("Lift Chart for Predicting Survival - Test Data Set") + 
  xlab("") + ylab("")
```

> Tip: `dplyr` and `sparklyr` both support windows functions, including `ntiles` and `cumsum`.


## AUC and accuracy

Though ROC curves are not available, Spark ML does have support for Area Under the ROC curve. This metric captures performance for specific cut-off values. The higher the AUC the better.

```{r auc}
# Function for calculating accuracy
calc_accuracy <- function(data, cutpoint = 0.5){
  data %>% 
    mutate(prediction = if_else(prediction > cutpoint, 1.0, 0.0)) %>%
    ml_classification_eval("prediction", "Survived", "accuracy")
}

# Calculate AUC and accuracy
perf_metrics <- data.frame(
  model = names(ml_score),
  AUC = 100 * sapply(ml_score, ml_binary_classification_eval, "Survived", "prediction"),
  Accuracy = 100 * sapply(ml_score, calc_accuracy),
  row.names = NULL, stringsAsFactors = FALSE)

# Plot results
gather(perf_metrics, metric, value, AUC, Accuracy) %>%
  ggplot(aes(reorder(model, value), value, fill = metric)) + 
  geom_bar(stat = "identity", position = "dodge") + 
  coord_flip() +
  xlab("") +
  ylab("Percent") +
  ggtitle("Performance Metrics")
```


## Feature importance

It is also interesting to compare the features that were identified by each model as being important predictors for survival. The logistic regression and tree models implement feature importance metrics. Sex, fare, and age are some of the most important features.

```{r importance, warning = FALSE}
# Initialize results
feature_importance <- data.frame()

# Calculate feature importance
for(i in c("Decision Tree", "Random Forest", "Gradient Boosted Trees")){
  feature_importance <- ml_tree_feature_importance(sc, ml_models[[i]]) %>%
    mutate(Model = i) %>%
    mutate(importance = as.numeric(levels(importance))[importance]) %>%
    mutate(feature = as.character(feature)) %>%
    rbind(feature_importance, .)
}

# Plot results
feature_importance %>%
  ggplot(aes(reorder(feature, importance), importance, fill = Model)) + 
  facet_wrap(~Model) +
  geom_bar(stat = "identity") + 
  coord_flip() +
  xlab("") +
  ggtitle("Feature Importance")
```

***

# Discuss

You can use `sparklyr` to run a variety of classifiers in Apache Spark. For the Titanic data, the best performing models were tree based models. Gradient boosted trees was one of the best models, but also had a much longer average run time than the other models. Random forests and decision trees both had good performance and fast run times.

While these models were run on a tiny data set in a local spark cluster, these methods will scale for analysis on data in a distributed Apache Spark cluster.